
\documentclass{uvamscse}

\input{program-listings}
\newcommand{\cmd}[1]{\texttt{$\backslash$#1}}

\usepackage[toc,acronym,nogroupskip]{glossaries}
\usepackage{url}

\usepackage{graphicx}
\graphicspath{ {figures/} }

\usepackage{booktabs}
\usepackage{multirow}
\usepackage{scrextend}

% \usepackage[subtle]{savetrees}

% ====== Listing Setup =======
\usepackage{color}
\usepackage{listings}
\usepackage{parcolumns}
\definecolor{javared}{rgb}{0.6,0,0} % for strings
\definecolor{javagreen}{rgb}{0.25,0.5,0.35} % comments
\definecolor{javapurple}{rgb}{0.5,0,0.35} % keywords
\definecolor{javadocblue}{rgb}{0.25,0.35,0.75} % javadoc
\definecolor{darkgray}{rgb}{.4,.4,.4}
 
\lstset{language=Java,
	basicstyle=\ttfamily,
	keywordstyle=\color{javapurple}\bfseries,
	stringstyle=\color{javared},
	commentstyle=\color{javagreen},
	morecomment=[s][\color{javadocblue}]{/**}{*/},
	keywordstyle={[2]\color{red}\bf},
	numbers=left,
	numberstyle=\tiny\color{black},
	stepnumber=1,
	numbersep=10pt,
	tabsize=2,
	showspaces=false,
	showstringspaces=false
}

\lstdefinelanguage{AspectJ}[]{Java}{
    morekeywords={declare, pointcut, aspect, before, around, after, returning, throwing, call, execution, this, target, args, within, withincode, get, set, initialization, preinitialization, staticinitialization, handler, adviceexecution, cflow, cflowbelow, if, proceed},
    moredelim=[is][\textcolor{darkgray}]{\%\%}{\%\%},
    moredelim=[il][\textcolor{darkgray}]{§§}
}

% ====== Glossaries setup =======
\makeglossaries
\newcommand{\ac}[1]{\gls{#1}}
\newcommand{\acs}[1]{\acrshort{#1}}
\newcommand{\acp}[1]{\glspl{#1}}

\input{acronyms}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Taming Aspects with Managed Data }
% \coverpic[100pt]{figures/terminal.png}
\subtitle{Master's Project in Software Engineering}
\date{Summer 2016}

\author{Theologos A. Zacharopoulos}
\authemail{theol.zacharopoulos@cwi.nl}
\supervisor{Tijs van der Storm}
\host{Centrum Wiskunde \& Informatica, \url{http://www.cwi.nl}}

\begin{document}

\abstract{
	The problem of crosscutting concerns results to code scattering and tangling.
	This in turn significantly affects the modularity of a system, since the separation of concerns principle is violated.
	Although Aspect Oriented Programming is created in order to solve this problem, several issues arise, the most important being the coupling between aspects and components.
	This leads to an evolution paradox in Aspect Oriented Software Development.

	In this thesis we implement managed data and use them in order to solve the problem of crosscutting concerns.
	Managed data is a data abstraction mechanism that allows the programmer to define data in addition to their manipulation mechanisms, something that is hard-coded in the programming languages.
	Defining crosscutting concerns in the manipulation mechanisms, namely data managers, results to a modular way of controlling aspects of data.
	Therefore, managed data can be used to solve the crosscutting concerns problem and avoid the coupling problem of Aspect Oriented Programming.

	Our work focuses on the implementation of managed data in Java, proving that managed data can be implemented in a static language.
	Moreover, we demonstrate that managed data can tame aspects by refactoring an existing program's crosscutting concerns.
	Finally, we provide an evaluation of our managed data refactoring in relation to an Aspect Oriented refactoring of the same use case application, in order to show how managed data can overcome Aspect Oriented Programming problems.
}

\maketitle

\printglossary[type=\acronymtype, title=Abbreviations]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Chapters
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\include{./chapters/Introduction}
\include{./chapters/Background}
\include{./chapters/Example}
\include{./chapters/Implementation}
\include{./chapters/AspectRefactoring}
\include{./chapters/Evaluation}
\include{./chapters/Conclusion}

% \chapter*{Acknowledgments}
% This is all started when my supervisor, Tijs van der Storm, provided us with the classic Aspect Oriented Programming paper during our software construction course of this Master's program.
% I was always thinking that crosscutting concerns is a problem of modularity and code cloning, scattering was annoying and I knew this was bad, but I never knew its name.
% But then the Aspect Oriented Programming paper explained it to me explicitly.
% First thing was to go to Tijs and ask him if AOP really works since I was not really convinced.
% How can no one use this technique if it really works?
% Tijs then explained to me that he and William R. Cook worked together on something different, that would solve that problem in a different way, the Managed Data.
% That day I left from his office with a list of papers discussing AOP and of course Managed Data.
% I've started implement it in Java right away in order to see if it actually works.
% Well, it worked, and I got very excited.
% This was the time I knew to do this.

% For this thesis, first I would like to express my sincere gratitude to Tijs van der Storm, my supervisor who introduced me to this idea.
% Tijs was a great inspiration for me from the first time.
% He helped me with every single problem I faced during this process.
% I will never forget the bootstraping process, Tijs had the patience to explain it to me a number of times, it was always mind-blowing, until I understand it.
% Additionally I want to thank Tijs because he was a real inspiration for me on his way of thinking and of course perfectionism. 
% He changed my way of programming and thinking about programming completely.

% Moreover, I would like to thank my colleagues and friends from CWI for the interesting and inspiring conversations.
% However, my biggest support was from Ifigeneia.
% The writing of this thesis would never be complete if she was not on my side.
% I express my thanks to Ifigeneia for her constant help, inspiration and the interest she showed on the subject during its completion.

\appendix
\include{./appendices/MObject}
\include{./appendices/SchemaLoading}
\include{./appendices/Migration}
\include{./appendices/ImplementingDataManager}
\include{./appendices/MetricsResults}

{%\tiny
  \bibliographystyle{alphaurl}
  \bibliography{thesis}
}

\end{document}
